OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA

### 2018-11-12 phil_paper/jmp_qje.tex
polishing jmp intro

%\usepackage[dvipsnames]{xcolor}
%\usepackage{cite}
\onehalfspacing
\usepackage{natbib}  
%\title{ Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila }
\title{ OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA}

@@

	Public utilities subsidize fixed connection fees with high marginal prices to provide access especially for the poor, but this policy can have the opposite effect when many households share connections.  Using data on 1.5 million water connections in Manila, I establish that informal sharing networks provide \input{tables/shr_use_from_neighbor}\unskip\% 
	of total access.  I structurally estimate household water demand across three sources: purchasing directly from the provider, sharing with a neighbor's tap, or buying from a small-scale vendor.  The model predicts that low fixed fees and high marginal prices decrease access by weakening incentives to share water.  In contrast, the optimal pricing policy features a high fixed fee and low marginal price, increases shared connections, and ensures nearly universal access to piped water compared to current prices in Manila.  This policy improves welfare by up to \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus or \input{tables/tpt_improvement_income_sharing_group}\unskip\% of household income.

@@

\end{abstract} 
{\bf JEL-Classification:} H23, L95, L98, O13, O17, O18 \\
\vspace{.1cm}
{\bf Keywords:} Water demand, non-linear pricing, informal sector, developing countries.

\section{Introduction}

Despite large investments in piped water throughout the developing world, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries for at least the past decade (World Bank [2015]).  Given health, time-savings, and other benefits from piped water, how can water utilities set prices to ensure access while covering costs?\footnote{\cite{gamper2010impact,cutler2005role}, and \cite{galiani2005water} find health improvements in both developed and developing settings while \cite{devoto2012happiness} find time savings and other utility benefits.}  A popular approach particularly in developing cities is to subsidize upfront connection fees with high marginal prices.\footnote{\cite{mcintosh2003asian} and \cite{komives2006distributional} from the World Bank and Asian Development Bank argue for low connection fees.  Using a randomized controlled trial from Morocco, \cite{devoto2012happiness} confirm that modest assistance with connection fees can lead to increases in new water connections.}  Over 70\% of developing cities also increase marginal prices with monthly usage so that large users cross-subsidize access and usage for smaller, poorer users.\footnote{This strategy is also commonly referred to as ``Increasing Block Tariffs'' or  ``Convex Pricing'' (\cite{borenstein2012redistributional}). \cite{hoque2013state} and \cite{boland2000political} provide overviews of the use of increasing marginal prices in the water sector in developing countries.}  Recent empirical papers in South Africa and C{\^o}te d'Ivoire predict gains in both access and total welfare from increasing marginal prices (\cite{szabo2015value,diakite2009proposal}).

Conventional pricing policies often rely on the assumption that each water connection serves a single household; however, many households in developing cities share with neighbors instead of purchasing their own connections.  For example, 70\% of African cities report frequent water sharing (\cite{keener2010provision}) ranging from fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) to drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  In Manila, I establish that 23\% of households access water through a neighbor's connection by connecting plumbing directly to the meter (\input{tables/access_pipes}\unskip\%), extending plastic pipes (\input{tables/access_single_tap}\unskip\%), or fetching water with containers (\input{tables/access_fetch}\unskip\%).  While previous studies have focused on specific suburbs, this paper is the first to my knowledge to measure the role of water sharing in extending access across an urban population.

Water sharing has important implications for the welfare impacts of common pricing policies.  Increasing marginal prices weaken incentives for households to extend water access to their neighbors through sharing (\cite{whittington1992possible}).  Fixed fee subsidies may have limited impacts on water access when sharing households already split connection fees with their neighbors.  Since low-income households are more likely to use water from their neighbors, increasing marginal prices and low fixed fees may combine to act as regressive taxes in the presence of sharing.

%  Shared connections record high levels of monthly consumption, which can expose them to high prices when marginal prices increase with usage . 
This paper analyzes the welfare impacts of pricing policies in the presence of informal water sharing.  Welfare impacts depend not only on heterogeneity in demand for water across the population, but also on costs associated with forming and maintaining water sharing relationships.  To measure these features, I merge administrative data covering 1.5 million water connections in Manila from 2010 to 2015 with detailed survey data on the number of households and people using each water connection.  With comprehensive geographic coverage of half of Manila and a variety of price variation, these data allow for a rich characterization of water demand.  Additional variation in water source choices identify the relative fixed and marginal costs associated with sharing water with a neighbor.  I find that despite relatively high marginal costs of sharing, shared connections allow the city to save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households at low marginal prices.  At the same time, informal sharing networks serve as resale markets for water, undermining the extent to which cities can redistribute surplus through increasing marginal prices.

To evaluate the impacts of different pricing policies, I build a model of household water demand that includes both initial source choices as well as monthly consumption choices.  This model expands on the demand estimation literature for public utilities that focuses almost exclusively on quantity choices (\cite{diakite2009proposal,mcrae2014infrastructure,olmstead2009reduced,szabo2015value}).  I specify a sequential game where households choose whether to connect individually, share with their neighbors, or purchase from a water vendor.  For sharing households, I develop a simple model of the water sharing contract that captures two key features of these relationships: (1) households face challenges in observing each other's demand and (2) households often split the monthly water bill evenly.  The model characterizes monthly consumption choices when households face non-linear budget constraints and imperfect control over their exact consumption levels.  This approach extends previous models of optimization over non-linear budget constraints by allowing households to anticipate unobserved shocks to their demand (\cite{moffitt1986econometrics,burtless1978effect}).

@@

Table~\ref{table:elasticity} presents connection-level demand elasticity estimates across terciles of (1) imputed income for the connection owner and (2) average monthly usage for the connection.  Elasticities exhibit little heterogeneity across imputed income levels, but substantial variation across usage levels with larger users appearing much less sensitive to prices than smaller users.  Since many low-income households share connections with neighbors, there is a weak correlation between income and usage-level, which helps account for the lack of heterogeneity in elasticity.\footnote{Noise in the income imputation process may also weaken any correlation between elasticity and income.}  Heterogeneity in elasticities according to monthly usage is consistent with moral hazard when households share a connection:  increasing the number of households per connection increases the water use for those connections while also exacerbating free-riding behavior.  Different water using activities across the water demand profile may also produce this heterogeneity in elasticities.  This heterogeneity provides a potential motivation for usage-based price discrimination.  For example, matching larger, less price-sensitive users with higher prices may allow for a more efficient allocation of water while still covering costs of production.
\begin{table}
\centering
\begin{threeparttable}
\caption{Demand Elasticities for Water Connections}\label{table:elasticity}
\input{tables/elasticity_estimates} 
\begin{tablenotes}
\item {\footnotesize Elasticities are calculated from 7,151 connections making 349,265 connection-months.  Income is imputed based on demographics, locations, and their interactions.  Standard errors are bootstrapped at the connection level.}
\end{tablenotes}
\end{threeparttable}
\end{table}
The demand estimation yields an average elasticity of \input{tables/mean_elasticity}\unskip, which can be interpreted as the percentage decrease in quantity due to a one percentage increase in price.  This estimate also closely matches reduced-form elasticity estimates from price-change quasi-experiment detailed in Appendix~\ref{appendix:ratechangeelasticity}.  This estimate is in range of the 0.98 estimate from \cite{szabo2015value} as well as similar studies in the developing world that find elasticities between 0.01 and 0.81 (\cite{diakite2009proposal}, \cite{strand2005water}). 
%is its inability to match zero consumption values observed in the data
\begin{table}
\caption{Model Performance}\label{table:modelperformance}
\input{tables/sample_prediction}
\end{table}
Table~\ref{table:modelperformance} compares predicted consumption from the model to observed consumption in the data.  This exercise finds negligible differences across average consumption and the variance of consumption for connections according to sharing status.  Closely matching observed patterns, this model improves on previous approaches to demand estimation likely by recovering a separate fixed preference term for each connection as well as rich heterogeneity in the variance terms (\cite{szabo2015value}).

>>>>

%\usepackage[dvipsnames]{xcolor}
%\usepackage{cite}
\onehalfspacing
\usepackage{natbib}  
%\title{ Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila }
\title{OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA}

@@

	Public utilities subsidize fixed connection fees with high marginal prices to provide access especially for the poor, but this policy can have the opposite effect when many households share connections.  Using data on 1.5 million water connections in Manila, I establish that informal sharing networks provide \input{tables/shr_use_from_neighbor}\unskip\% 
	of total access.  I structurally estimate household water demand across three sources: purchasing directly from the provider, sharing with a neighbor's tap, or buying from a small-scale vendor.  The model predicts that low fixed fees and high marginal prices decrease access by weakening incentives to share water.  In contrast, the optimal pricing policy features a high fixed fee and low marginal price, increases shared connections, and ensures nearly universal access to piped water compared to current prices in Manila.

	%This policy improves welfare by up to \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus.

	% or \input{tables/tpt_improvement_income_sharing_group}\unskip\% of household income

@@

\end{abstract} 
\vspace{1cm}
{\bf JEL-Classification:} H23, L95, L98, O13, O17, O18 \\
\vspace{.1cm}
{\bf Keywords:} Water demand, non-linear pricing, informal sector, developing countries.

\section{Introduction}
 % \footnote{\cite{laffont2005regulation} emphasizes high costs of public funds in developing countries (with a deadweight loss of 2.5 for the Philippines) due to limited ability to enter long-term contracts with government, credit constraints, inefficient taxation, and corruption.}
 % To ensure widespread access while covering costs, a 
Improving access to piped water generates substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  However, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, government subsidies are often limited by high costs of public funds and risk of regulatory capture in these countries (\cite{laffont2005regulation}).  Instead, prices must often be set to fully recover costs.  A popular solution is to subsidize upfront connection fees with high marginal prices for monthly usage.\footnote{\cite{mcintosh2003asian} and \cite{komives2006distributional} from the World Bank and Asian Development Bank argue for low connection fees.  Using a randomized controlled trial from Morocco, \cite{devoto2012happiness} confirm that modest assistance with connection fees can lead to increases in new water connections.}  Over 70\% of developing cities also increase marginal prices with monthly usage so that large users cross-subsidize access and usage for smaller, poorer users.\footnote{This strategy is also commonly referred to as ``Increasing Block Tariffs'' or  ``Convex Pricing'' (\cite{borenstein2012redistributional}). \cite{hoque2013state} and \cite{boland2000political} provide overviews of the use of increasing marginal prices in the water sector in developing countries.  The statistic comes from author calculations with data kindly provided by \cite{hoque2013state}.}  These pricing approaches have been shown to work well when each water connection serves a single household.  Recent empirical papers in South Africa and C{\^o}te d'Ivoire predict gains in both access and total welfare from increasing marginal prices (\cite{szabo2015value,diakite2009proposal}).

However, many households in developing cities share connections with their neighbors, which can have large implications for water pricing.  Land-tenure issues, credit-constraints, and administrative costs often prevent disadvantaged households from purchasing their own connections (\cite{devoto2012happiness}).  Also, many cities maintain high upfront fees despite policy recommendations otherwise.  As a result, 70\% of African utilities report frequent water sharing between households (\cite{keener2010provision}) ranging from fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) to drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of informal sharing, previous research has not yet examined the importance of informal sharing for optimal pricing policy.

This paper analyzes the welfare effects of pricing policies in the presence of informal water sharing.  I build a model of household water demand that includes both (1) the upfront choice of whether to purchase a connection, share with a neighbor, or use non-piped sources as well as (2) monthly consumption choices.  I also collect new survey evidence on informal sharing contracts to inform several features of the model design.  The structural estimation leverages administrative data from 1.5 million water connections in Manila, which link to survey data on households owning and sharing connections.

These estimates allow me to solve a standard regulator pricing problem: setting prices to cover both fixed and marginal costs while maximizing consumer surplus.  I also allow the regulator to place extra weight on surplus for smaller, poorer users.  In the presence of sharing, the standard approach --- low fixed fees and high marginal prices --- excludes many small-users from piped water by reducing incentives to share connections.  Instead, these households substitute toward non-piped water sources, which have lower marginal prices, but varying quality.\footnote{Common non-piped water sources in Manila include deepwells, tanker trucks, and water refilling stations.  The model collapses all non-piped sources into a single ``vendor'' option.}  The opposite approach --- high fixed fees and low marginal prices --- improves both access and welfare even for smaller users.  High fixed fees are paid in full by households who connect individually and are split between sharing households.  Informal sharing arrangements allow small-users to access piped water from their neighbors at low marginal prices.  Shared connections also let the city save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households.

%By using from their neighbors, smaller-users also benefit from low marginal prices and avoid sourcing from non-piped water.  
% Non-linear pricing helps policymakers achieve both efficiency (by targeting more inelastic households with higher prices) as well as equity (by targeting poorer households with lower prices).
% excluding many smaller-users from piped water.\footnote{In the model, households can substitute to a vendor option that encompasses all non-piped sources of water, which commonly include deepwells, tanker trucks, and water refilling stations.}  
%The model is designed to match several features from new survey evidence on sharing contracts that I fielded as part of this study.

%In the context of Manila, I find that the most disadvantaged households are also most likely to fetch water from their neighbors.
%In Manila, I establish that 23\% of households access water through a neighbor's connection by connecting plumbing directly to the meter (\input{tables/access_pipes}\unskip\%), extending plastic pipes (\input{tables/access_single_tap}\unskip\%), or fetching water with containers (\input{tables/access_fetch}\unskip\%).

Manila

To measure these features, I merge administrative data covering 1.5 million water connections in Manila from 2010 to 2015 with detailed survey data on the number of households and people using each water connection.

With comprehensive geographic coverage of half of Manila and a variety of price variation, these data allow for a rich characterization of water demand.  Additional variation in water source choices identify the relative fixed and marginal costs associated with sharing water with a neighbor.

I find that shared connections allow the city to save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households at low marginal prices.

%At the same time, informal sharing networks serve as resale markets for water, undermining the extent to which cities can redistribute surplus through increasing marginal prices.

%  Shared connections record high levels of monthly consumption, which can expose them to high prices when marginal prices increase with usage .

Welfare impacts depend not only on heterogeneity in demand for water across the population, but also on costs associated with forming and maintaining water sharing relationships.  [DONT LIKE THIS SENTENCE ANYMORE]

To evaluate the impacts of different pricing policies, I build a model of household water demand that includes both initial source choices as well as monthly consumption choices.

This model expands on the demand estimation literature for public utilities that focuses almost exclusively on quantity choices (\cite{diakite2009proposal,mcrae2014infrastructure,olmstead2009reduced,szabo2015value}).  I specify a sequential game where households choose whether to connect individually, share with their neighbors, or purchase from a water vendor.  For sharing households, I develop a simple model of the water sharing contract that captures two key features of these relationships: (1) households face challenges in observing each other's demand and (2) households often split the monthly water bill evenly.  The model characterizes monthly consumption choices when households face non-linear budget constraints and imperfect control over their exact consumption levels.  This approach extends previous models of optimization over non-linear budget constraints by allowing households to anticipate unobserved shocks to their demand (\cite{moffitt1986econometrics,burtless1978effect}).

@@

Table~\ref{table:elasticity} presents connection-level demand elasticity estimates across terciles of (1) imputed income for the connection owner and (2) average monthly usage for the connection.  Elasticities exhibit little heterogeneity across imputed income levels, but substantial variation across usage levels with larger users appearing much less sensitive to prices than smaller users.  Since many low-income households share connections with neighbors, there is a weak correlation between income and usage-level, which helps account for the lack of heterogeneity in elasticity.\footnote{Noise in the income imputation process may also weaken any correlation between elasticity and income.}  Heterogeneity in elasticities according to monthly usage is consistent with moral hazard when households share a connection:  increasing the number of households per connection increases the water use for those connections while also exacerbating free-riding behavior.  Different water using activities across the water demand profile may also produce this heterogeneity in elasticities.  This heterogeneity provides a potential motivation for usage-based price discrimination.  For example, matching larger, less price-sensitive users with higher prices may allow for a more efficient allocation of water while still covering costs of production.
\begin{table}
\centering
\begin{threeparttable}
\caption{Demand Elasticities for Water Connections}\label{table:elasticity}
\input{tables/elasticity_estimates} 
\begin{tablenotes}
\item {\footnotesize Elasticities are calculated from 7,151 connections making 349,265 connection-months.  Income is imputed based on demographics, locations, and their interactions.  Standard errors are bootstrapped at the connection level.}
\end{tablenotes}
\end{threeparttable}
\end{table}
The demand estimation yields an average elasticity of \input{tables/mean_elasticity}\unskip, which can be interpreted as the percentage decrease in quantity due to a one percentage increase in price.  This estimate also closely matches reduced-form elasticity estimates from price-change quasi-experiment detailed in Appendix~\ref{appendix:ratechangeelasticity}.  This estimate is in range of the 0.98 estimate from \cite{szabo2015value} as well as similar studies in the developing world that find elasticities between 0.01 and 0.81 (\cite{diakite2009proposal}, \cite{strand2005water}). 
%is its inability to match zero consumption values observed in the data
\begin{table}
\caption{Model Performance}\label{table:modelperformance}
\centering
\input{tables/sample_prediction}
\end{table}
Table~\ref{table:modelperformance} compares predicted consumption from the model to observed consumption in the data.  This exercise finds negligible differences across average consumption and the variance of consumption for connections according to sharing status.  Closely matching observed patterns, this model improves on previous approaches to demand estimation likely by recovering a separate fixed preference term for each connection as well as rich heterogeneity in the variance terms (\cite{szabo2015value}).
